Analytic Occultation Light Curves in Reflected Light

### 2020-04-02 tex/ms.tex
better problem setup

% Shorthand for this paper
\usepackage{xifthen}
\newcommand{\BF}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\BS}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\dd}{\ensuremath{\mathrm{d}}}
\newcommand{\STARRYQUADPOINTS}{100\xspace}
\newcommand{\bkappa}{\BS{\kappa}}
\newcommand{\blambda}{\BS{\lambda}}
\newcommand{\bxi}{\BS{\xi}}
\newcommand{\vmax}{{v_\mathrm{max}}}
\newcommand{\kapint}[1]{%
\sum_{i = 0}^{\frac{N - 1}{2}}
\int\limits_{\frac{1}{2}\kappa_{2i}}^{\frac{1}{2}\kappa_{2i+1}}
#1
\dd\varphi
}
\newcommand{\lamint}[1]{%
\sum_{i = 0}^{\frac{N - 1}{2}}
\int\limits_{\frac{1}{2}\lambda_{2i}}^{\frac{1}{2}\lambda_{2i+1}}
#1
\dd\varphi
}
\newcommand{\coshalfkap}[1][]{%
\ifthenelse{%
    \equal{#1}{}
}{
    \cos\left(\frac{\bkappa}{2}\right)
}{
    \cos^{#1}\left(\frac{\bkappa}{2}\right)
}}
\newcommand{\sinhalfkap}[1][]{%
\ifthenelse{%
    \equal{#1}{}
}{
    \sin\left(\frac{\bkappa}{2}\right)
}{
    \sin^{#1}\left(\frac{\bkappa}{2}\right)
}}
\newcommand{\DE}{\Delta \BS{E}(k^2, \bkappa)}
\newcommand{\DF}{\Delta \BS{F}(k^2, \bkappa)}
\newcommand{\sgn}{{\mathrm{sgn}}}
\newcommand{\xhat}{\ensuremath{\mathbf{\hat{x}}}\xspace}
\newcommand{\yhat}{\ensuremath{\mathbf{\hat{y}}}\xspace}
\newcommand{\zhat}{\ensuremath{\mathbf{\hat{z}}}\xspace}
\newcommand{\sT}{\ensuremath{\BF{s}^\top}}
\newcommand{\bg}{\ensuremath{\tilde{\BF{g}}}}
\newcommand{\bp}{\ensuremath{\tilde{\BF{p}}}}

@@

% Title
\title{Occultation Light Curves in Reflected Light}

@@

%
\section{The Problem}
\label{sec:the-problem}
%
\subsection{Overview}
\label{sec:overview}
%
Without loss of generality, assume the body whose flux we wish to compute
has radius unity and sits at the origin of a right-handed Cartesian coordinate
system with the $x$-axis pointing to the right on the sky, the $y$-axis
pointing up on the sky, and the $z$-axis pointing out of the sky
toward the observer; denote this frame by $\mathcal{F}_0$. In $\mathcal{F}_0$,
the surface intensity field of the body is described by the spherical harmonic
coefficient vector $\mathbf{y}$. Following \citet{Luger2019}, we may compute
the total visible flux $f$ from this body as
%
\begin{align}
    \label{eq:sTARRy}
    f = \sT \BF{A} \BF{R}' \BF{R} \BF{y}
    \quad,
\end{align}
%
where, from right to left, $\BF{R} = \BF{R}(i, \lambda, \vartheta)$
is a Wigner rotation matrix that
rotates $\bvec{y}$ to the sky frame $\mathcal{F}_\mathrm{sky}$
given the body's inclination $i$, obliquity $\lambda$, and
rotational phase $\vartheta$,
%
$\BF{R}'$ rotates the body on the plane
of the sky into the frame $\mathcal{F}_\mathrm{int}$ in which the
integration is actually performed,
%
$\BF{A}$ is the change-of-basis matrix from spherical harmonics
to the basis in which the integrals are computed
%,
and $\sT$ is the \emph{solution vector}, the integral of each
of the basis functions over
the projected visible disk of the body (which may be partially occulted by
another spherical body). For more details, see Appendix~\ref{app:starry}.

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=0.35\linewidth]{figures/illum.pdf}
        \oscaption{illum}{%
            Illumination geometry for the unocculted case.
            A point source at Cartesian coordinates
            $(0, y_s, z_s)$ illuminates a uniform,
            Lambertian, spherical body of radius unity. The observed
            radiance falls off as the cosine of the angle $\phi$ between the
            source and the surface normal. Beyond the terminator
            (defined by $\phi = \frac{\pi}{2}$), the radiance is zero. In
            projection, the terminator is a segment of an ellipse of semi-major
            axis unity and semi-minor axis
            $b = -\nicefrac{z_s}{\sqrt{y_s^2 + z_s^2}}$.
            \label{fig:illum}
        }
    \end{centering}
\end{figure}

The derivations in \citet{Luger2019} assume that the coefficient vector
$\mathbf{y}$ describes the \emph{emissivity} of the body, which (in the
absence of limb darkening) is assumed to be Lambertian, i.e., all points on the
surface emit equally in all directions.
Here, we derive the solution to the occultation problem for Lambertian
reflectance, in which case the vector $\mathbf{y}$ describes the
surface \emph{reflectivity} (or \emph{Bond albedo}). Specifically, we assume
the body is illuminated by a point-like source and the radiance at any point
is proportional to the cosine of the angle $\phi$ between the incident light
and the surface normal. Points for which $\phi \ge \frac{\pi}{2}$ are
unilluminated and therefore have radiance of zero; see Figure~\ref{fig:illum}.

In \citet{Luger2019}, the crux of the problem was to compute the terms in the
vector $\sT$, which are surface integrals over
the region of the occulted body that is visible to
the observer (Figure 2 in that paper). Here, these integrals are significantly
more difficult, because they involve not only the intersection of the
limbs of the two bodies, but also intersections with the elliptical
day/night terminator of the occulted body.

>>>>

% Shorthand for this paper
\usepackage{xifthen}
\newcommand{\BF}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\BS}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\dd}{\ensuremath{\mathrm{d}}}
\newcommand{\STARRYQUADPOINTS}{100\xspace}
\newcommand{\bkappa}{\BS{\kappa}}
\newcommand{\blambda}{\BS{\lambda}}
\newcommand{\bxi}{\BS{\xi}}
\newcommand{\vmax}{{v_\mathrm{max}}}
\newcommand{\kapint}[1]{%
\sum_{i = 0}^{\frac{N - 1}{2}}
\int\limits_{\frac{1}{2}\kappa_{2i}}^{\frac{1}{2}\kappa_{2i+1}}
#1
\dd\varphi
}
\newcommand{\lamint}[1]{%
\sum_{i = 0}^{\frac{N - 1}{2}}
\int\limits_{\frac{1}{2}\lambda_{2i}}^{\frac{1}{2}\lambda_{2i+1}}
#1
\dd\varphi
}
\newcommand{\coshalfkap}[1][]{%
\ifthenelse{%
    \equal{#1}{}
}{
    \cos\left(\frac{\bkappa}{2}\right)
}{
    \cos^{#1}\left(\frac{\bkappa}{2}\right)
}}
\newcommand{\sinhalfkap}[1][]{%
\ifthenelse{%
    \equal{#1}{}
}{
    \sin\left(\frac{\bkappa}{2}\right)
}{
    \sin^{#1}\left(\frac{\bkappa}{2}\right)
}}
\newcommand{\DE}{\Delta \BS{E}(k^2, \bkappa)}
\newcommand{\DF}{\Delta \BS{F}(k^2, \bkappa)}
\newcommand{\sgn}{{\mathrm{sgn}}}
\newcommand{\xhat}{\ensuremath{\mathbf{\hat{x}}}\xspace}
\newcommand{\yhat}{\ensuremath{\mathbf{\hat{y}}}\xspace}
\newcommand{\zhat}{\ensuremath{\mathbf{\hat{z}}}\xspace}
\newcommand{\sT}{\ensuremath{\BF{s}^\top}}
\newcommand{\rT}{\ensuremath{\BF{r}^\top}}
\newcommand{\bg}{\ensuremath{\tilde{\BF{g}}}}
\newcommand{\bp}{\ensuremath{\tilde{\BF{p}}}}
\newcommand{\by}{\ensuremath{\tilde{\BF{y}}}}

@@

% Title
\title{Analytic Occultation Light Curves in Reflected Light}

@@

%
\section{The Problem}
\label{sec:the-problem}
%
\subsection{Review of the \starry algorithm in emitted light}
\label{sec:starry-review}
%
Without loss of generality, assume the body whose flux we wish to compute
has radius unity and sits at the origin of a right-handed Cartesian coordinate
system with the $x$-axis pointing to the right on the sky, the $y$-axis
pointing up on the sky, and the $z$-axis pointing out of the sky
toward the observer; denote this frame by $\mathcal{F}_0$. In $\mathcal{F}_0$,
the surface (emitted) intensity field of the body is described by the
spherical harmonic coefficient vector $\mathbf{y}$. Following
\citet{Luger2019}, when no occultor is present, we may compute the total
visible flux $f$ from this body as
%
\begin{align}
    \label{eq:rTA1Ry}
    f = \rT \BF{A_1} \BF{R} \BF{y}
    \quad,
\end{align}
%
where, from right to left, $\BF{R} = \BF{R}(i, \lambda, \vartheta)$
is a Wigner rotation matrix that rotates $\bvec{y}$ to the sky frame
$\mathcal{F}_\mathrm{sky}$ given the body's inclination $i$, obliquity
$\lambda$, and rotational phase $\vartheta$,
%
$\BF{A_1}$ is the change-of-basis matrix from the spherical harmonic
basis $\by$ to the \emph{polynomial basis} $\bp$ in which the integrals
are computed,
%
and $\rT$ is the vector of solutions to the integral over
the projected visible disk of the body for each term in $\bp$.
%
If an occultor is present, we instead compute the visible flux from
%
\begin{align}
    \label{eq:sTARRy}
    f = \sT \BF{A} \BF{R}' \BF{R} \BF{y}
    \quad,
\end{align}
%
where, as before, $\BF{R}$ rotates the body from $\mathcal{F}_0$
to $\mathcal{F}_\mathrm{sky}$,
%
$\BF{R}'$ rotates the body on the plane
of the sky into the frame $\mathcal{F}_\mathrm{greens}$ in which the
integration is actually performed,
%
$\BF{A}$ is the change-of-basis matrix from $\by$
to the \emph{Green's basis} $\bg$ in which the integrals are computed,
%
and $\sT$ is the vector of solutions to the integral over
the projected visible disk of the body for each term in $\bg$.
For more details, see Appendix~\ref{app:starry} and \citet{Luger2019}.

\subsection{Adapting the algorithm to the reflected light case}
\label{sec:starry-review}
%
In order to compute light curves in reflected light, we must make two
modifications to the \starry algorithm. First,
the expressions above assume that the coefficient vector
$\mathbf{y}$ describes the \emph{emissivity} of the body, which (in the
absence of limb darkening) is assumed to be Lambertian, i.e., all points on the
surface emit equally in all directions.
Here, we wish to derive the solution for the flux in the case of Lambertian
reflectance, in which case the vector $\mathbf{y}$ is taken to describe the
surface \emph{reflectivity} (or \emph{Bond albedo}).

Second, we must explicitly model the illumination of the body. We assume the
body is illuminated by a point-like source, in which case the radiance at any
point on the surface is proportional to the cosine of the angle $\phi$ between
the incident light and the surface normal. Points for which
$\phi \ge \frac{\pi}{2}$ are unilluminated and therefore have radiance of zero.
If the point-like illumination source is placed at sky coordinates
$(x_s, y_s, z_s)$, the day/night terminator on the body is a half-ellipse
of semi-major axis unity that is fully described by its semi-minor axis,
    \label{eq:b}
    b = -\frac{z_s}{\sqrt{x_s^2 + y_s^2 + z_s^2}}
    \quad,
and the angle by which its semi-major axis is rotated away from the
$+x$-axis,
    \label{eq:theta}
    \theta = \mathrm{arctan2}(y_s, x_s)
Given this formulation, and assuming that
$x_s^2 + y_s^2 + z_s^2 \gg 1$,
it is straightforward to show that the illumination
$I$ at a point $(x, y)$ on the projected disk of the body is given by
the polynomial
    \frac{I(b, \theta; x, y)}{I_0} &=
    -b_c\sin\theta x + b_c\cos\theta y - bz(x, y)
where $b_c = \sqrt{1 - b^2}$, $z(x, y) = \sqrt{1 - x^2 - y^2}$, and
    I_0 = \frac{L}{4\pi\left(x_s^2 + y_s^2 + z_s^2\right)}
is the peak illumination (at the substellar point,
assuming the illumination source is a star), with $L$ equal to the luminosity
of the source.
Since the illumination is just a polynomial in $x$, $y$, and $z(x, y)$, we
can express it as a vector $\BF{i}(b, \theta)$ in the polynomial basis $\bp$.
Recalling the structure of the basis (c.f. Appendix~\ref{app:starry}),
    \bp & =
    \begin{pmatrix}
        1   &
        x   & z  & y  &
        x^2 & xz & xy & yz & y^2 &
        \cdot\cdot\cdot
    \end{pmatrix}^\mathsf{T}
    \quad,
we may write
    \BF{i}(b, \theta) & =
    I_0
    \begin{pmatrix}
        0              \\
        -b_c\sin\theta \\
        -b             \\
        b_c\cos\theta
    \end{pmatrix}
This fact allows us to construct a linear operator $\BF{I}$ to weight a map
vector in the polynomial basis by the illumination profile.
If we think about how each of the terms in $\bp$ transforms under $\BF{I}$,
\begin{minipage}{0.22\linewidth}
        \begin{pmatrix}
            1 \\
            0 \\
            0 \\
            0
        \end{pmatrix}
         & \BS{\rightarrow}
        \begin{pmatrix}
            \bvec{i}_0 \\ %1
            \bvec{i}_1 \\ %x
            \bvec{i}_2 \\ %z
            \bvec{i}_3 \\ %y
            0          \\ %x^2
            0          \\ %xz
            0          \\ %xy
            0          \\ %yz
            0             %y^2
        \end{pmatrix}
\begin{minipage}{0.22\linewidth}
        \begin{pmatrix}
            0 \\
            1 \\
            0 \\
            0
        \end{pmatrix}
         & \BS{\rightarrow}
        \begin{pmatrix}
            0          \\ %1
            \bvec{i}_0 \\ %x
            0          \\ %z
            0          \\ %y
            \bvec{i}_1 \\ %x^2
            \bvec{i}_2 \\ %xz
            \bvec{i}_3 \\ %xy
            0          \\ %yz
            0    %y^2
        \end{pmatrix}
\begin{minipage}{0.22\linewidth}
        \begin{pmatrix}
            0 \\
            0 \\
            1 \\
            0
        \end{pmatrix}
         & \BS{\rightarrow}
        \begin{pmatrix}
            \bvec{i}_2  \\ %1
            0           \\ %x
            \bvec{i}_0  \\ %z
            0           \\ %y
            -\bvec{i}_2 \\ %x^2
            \bvec{i}_1  \\ %xz
            0           \\ %xy
            \bvec{i}_3  \\ %yz
            -\bvec{i}_2    %y^2
        \end{pmatrix}
\begin{minipage}{0.22\linewidth}
        \begin{pmatrix}
            0 \\
            0 \\
            0 \\
            1
        \end{pmatrix}
         & \BS{\rightarrow}
        \begin{pmatrix}
            0          \\ %1
            0          \\ %x
            0          \\ %z
            \bvec{i}_0 \\ %y
            0          \\ %x^2
            0          \\ %xz
            \bvec{i}_1 \\ %xy
            \bvec{i}_2 \\ %yz
            \bvec{i}_3    %y^2
        \end{pmatrix}
\begin{minipage}{0.05\linewidth}
we can compose $\BF{I}$ out of these column vectors:
    \BF{I}(b, \theta) & =
    I_0
    \begin{pmatrix}
        0              & 0              & -b             & 0              & \cdots \\
        -b_c\sin\theta & 0              & 0              & 0              & \cdots \\
        -b             & 0              & 0              & 0              & \cdots \\
        b_c\cos\theta  & 0              & 0              & 0              & \cdots \\
        0              & -b_c\sin\theta & b              & 0              & \cdots \\
        0              & -b             & -b_c\sin\theta & 0              & \cdots \\
        0              & b_c\cos\theta  & 0              & -b_c\sin\theta & \cdots \\
        0              & 0              & b_c\cos\theta  & -b             & \cdots \\
        0              & 0              & b              & b_c\cos\theta  & \cdots \\
        \vdots         & \vdots         & \vdots         & \vdots         & \ddots
    \end{pmatrix}
where the dimensions of the matrix are $\big((l + 2)^2, (l + 1)^2\big)$, where
$l$ is the spherical harmonic degree of the map (this operator raises the
degree of the map by one).

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/frames.pdf}
        \oscaption{frames}{%
            How \starry computes the flux from a body in reflected light,
            tracking each of the linear transformations from the input map
            (far right) to the output (far left). The label below each map
            denotes the reference frame, while the label above
            each map denotes the basis in which the map is represented.
            Arrows indicate linear operations and are labeled accordingly.
            The upper branch corresponds to the unocculted case, while the
            lower branch corresponds to the case where an occultor is present.
            See text for details.
            \label{fig:frames}
        }
    \end{centering}
\end{figure}

We may now re-write Equations~(\ref{eq:rTA1Ry}) and (\ref{eq:sTARRy}) to
account for this illumination transformation. The flux outside of occultation
is now given by
%
\begin{align}
    \label{eq:rTIA1RRy}
    f = \rT \BF{I} \BF{A_1} \BF{R}'' \BF{R} \BF{y}
    \quad,
\end{align}
%
and the flux during an occultation by
%
\begin{align}
    \label{eq:sTA2IA1RRy}
    f = \sT \BF{A_2} \BF{I} \BF{A_1} \BF{R}' \BF{R} \BF{y}
    \quad.
\end{align}
%
Two notes about these equations. First, we introduced a new rotation matrix
$\BF{R}''$ in Equation~(\ref{eq:rTIA1RRy}), which rotates the problem so that
the semi-major axis of the terminator is aligned with the $x$-axis; as will
become clear in \S\ref{sec:case-0} below, this greatly simplifies the
integration step. Second, in the derivation of Equation~(\ref{eq:sTA2IA1RRy})
we used the fact that $\BF{A} = \BF{A_2} \BF{A_1}$
\citep[c.f. Equation~14 in][]{Luger2019}, where $\BF{A_1}$ transforms from
the spherical harmonic basis $\by$ to the polynomial basis $\bp$, and
$\BF{A_2}$ transforms from $\bp$ to the Green's basis $\bg$.

Figure~\ref{fig:frames} summarizes the transformations involved in the two
equations above. Starting on the right with a map vector $\BF{y}$ in
the spherical harmonic basis $\by$, defined in some observer-independent frame
$\mathcal{F}_0$, we first rotate it via $\BF{R}$ to the sky frame
$\mathcal{F}_\mathrm{sky}$, in which orientation the body is viewed by the
observer. If there is no occultation (upper branch of the figure), we then
rotate the map once more via $\BF{R}''$ to the integration frame
$\mathcal{F}_\mathrm{int}$, in which the terminator is parallel to the
$x$-axis. We then apply $\BF{A_1}$ to change basis to $\bp$, apply the
illumination transform $\BF{I}$, and finally dot in the solutions to the
surface integrals $\rT$. If, on the other hand, an occultor is present,
we rotate the map from $\mathcal{F}_\mathrm{sky}$ via $\BF{R}'$ to the frame
$\mathcal{F}_\mathrm{greens}$, in which the occultor lies along the
$+y$-axis. We then apply $\BF{A_1}$ to change basis to $\bp$ and $\BF{I}$
to weight the map by the illumination as above. Finally, we change basis
via $\BF{A_2}$ to the Green's basis, in which we compute and dot the
integrals $\sT$.

In \citet{Luger2019}, the crux of the problem was to compute the terms in the
vectors $\sT$ and $\rT$, which are surface integrals over
the region of the body that is visible to the observer
(see Figure 2 in that paper). Here, these integrals are significantly
more difficult, because they involve not only integration along the
limb of the body and of a possible occultor, but also along the
elliptical day/night terminator.
